Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-01-19 vertical-thermal-review.Rmd
small edits

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig. `r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves -- which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004]. Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see sections 2.2, 4.1).

@@

Under hot and dry conditions, leaves face a trade-off between $T_{leaf}$ regulation and water conservation [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
With adequate water, high $\lambda E$ can facilitate heat dissipation, particularly for leaves with smaller size (higher $g_b$) and larger $g_s$ [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
However, when leaf water demand (determined by VPD and stomatal opening) exceeds the rate of supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
Therefore, at high solar radiation loads, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at maximum $g_s$, but during $g_s$ limitation, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].

@@

Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both predicted by optamization theory [@fieldAllocatingLeafNitrogen1983; @hiroseMaximizingDailyCanopy1987] and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under conditions of high light and sufficient water (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).

@@

Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] and photosynthesis.
Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimate2022].
In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimate2022], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015]. 
This points to a dominant role of vertical profiles in the biophysical environment (Fig. `r fig_NEON_vertical`) in shaping tree growth rates within forests.

@@

Future forest dynamics remain one of the largest sources of uncertainty in Earth system model projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006].
Dynamic global vegetation models (DGVMs), which comprise the land surface models in Earth system models, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004]. 
DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005]. 
This simplification is computationally more efficient, although it comes at a cost of a lack of representation of important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001]. 
The computational middle-ground to representing vertical structure in DVGMs lies in cohort-based models, which represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type [@fisherVegetationDemographicsEarth2018].
These sit between the oversimplified vegetation dynamics in big-leaf models, which do not represent any vertical stratification, and individual-based models, whose computational expense prevents them from being integrated into Earth system models.

>>>>

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig. `r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020]. 
In combination with $T_{air}$ and $T_{leaf}$, RH determines leaf-to-air vapor pressure deficit (VPD) -- the driving force of water loss from leaves -- which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004]. Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see sections 2.2, 4.1).

@@

Under hot and dry conditions, leaves face a trade-off between $T_{leaf}$ regulation and water conservation [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
With adequate water, high $\lambda E$ can facilitate heat dissipation, particularly for leaves with smaller size (higher $g_b$) [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
However, when leaf water demand (determined by VPD and stomatal opening) exceeds the rate of supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
Therefore, at high solar radiation loads, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at maximum $g_s$, but during $g_s$ limitation, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].

@@

Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both predicted by optamization theory [@fieldAllocatingLeafNitrogen1983; @hiroseMaximizingDailyCanopy1987] and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under conditions of high light and sufficient water (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).

@@

Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] and photosynthesis.
Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimate2022].
In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimate2022], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015]. 
This points to a dominant role of vertical profiles in the biophysical environment, particularly light (Fig. `r fig_NEON_vertical`), in shaping tree growth rates within forests.

@@

Future forest dynamics remain one of the largest sources of uncertainty in Earth system model projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006].
Dynamic global vegetation models (DGVMs), which comprise the land surface models in Earth system models, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004]. 
DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005]. 
This simplification is computationally more efficient, although it comes at a cost of a lack of representation of important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001]. 
The computational middle-ground to representing vertical structure in DVGMs lies in cohort-based models, which represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type [@fisherVegetationDemographicsEarth2018].
These sit between the oversimplified vegetation dynamics in big-leaf models, which do not represent any vertical stratification, and individual-based models, whose computational expense prevents them from being integrated into Earth system models.
